Estimating Daily Bicycle Counts in Seattle from Seasonal and Weather Factors

### 2014-12-09 article/article.tex
Readability tweaks to layout, a few assorted edits

% Layout options
\chapterstyle{article}
\setulmarginsandblock{.5in}{.75in}{*}
\setlrmarginsandblock{.5in}{*}{*}
\checkandfixthelayout
\raggedbottom

@@

Bicycling offers a variety of social benefits that include the health
of those choosing to bicycle regularly, as well as more broadly a
society-wide increase in mobility as people shift from congestion
causing modes to bicycling. As such, bicycling is being promoted by
policy makers and urban planners as an increasingly important element
of urban transportation. In order to develop policy and
infrastructural improvements to induce more bicycling, a robust
understanding of factors related to bicycling is required.

This article contributes to that understanding by focusing on the
substantive effects of weather and seasonality on the number of
bicyclists observed on a given day. Specifically, we hypothesize that
factors associated with inclement weather (low temperature and
precipitation) will be inversely correlated with bicycle counts. We
similarly expect to see an additional decrease in counts associated
with physical and social / institutional aspects of season. That is, in seasons
where the days get shorter, we expect fewer cyclists due to a kind of
decision making inertia. We also expect fewer cyclists outside
the ``school season.''

@@

\newfixedcaption{\freetabcaption}{table}
\begin{centering}
\begin{scriptsize}
\freetabcaption{Summary of findings from literature review}
\begin{ctabular}{p{.6in}p{1.25in}p{1.25in}}
  \toprule
  Source & Variable(s) identified & Methods \\
  \midrule
  \cite{Fields:2012aa} & Average daily temperature (F); Total weekly precipitation (inches) & Identify patterns through scatter plots; No explicit model is established. \\
  \cite{Gallop:2012aa} & Temperature, Relative humidity, wind speed, clearness,fog, precipitation & Use ARIMA to account for serial correlation patterns \\
  \cite{Griswold:2011aa} & Nearby population and employment density, proximity to downtown/freeway, age, education level, income, et. & Log linear ordinary least squares regression is used to estimate a bicycle count model \\
  \cite{Helbich:2014aa} & Daily maiximum air temperature (C), daily average wind speed (m/s), daily precipitation(mm) & Place-specific associations of weather conditions are explored through geographically weighted logit models \\
  \cite{Hunt:2007aa} & Descriptive variables indicating lane use, secured parking, level of experience, etc. & Logistic model of cycling-related choices \\
  \cite{Jones:2010aa} & Length of bicycle network, employment density, population density & Standard ordinary least squares regression \\
  \cite{Lewin:2011aa} & Max temp (F), rain flag, snow flag, weekend flag, over 90 flag & Standard linear regression model \\
  \cite{McCahil:2008aa} & logarithmic choice measure, population density, worker density & A new space syntax theory is used to evaluate and predict the bicycle volume throughout a network \\
  \cite{Miranda-Moreno:2011aa} & Temperature (C), humidity (\%), rain presence (0/1), rain in prev. 3hrs (0/1), warm \& humid (0/1), morning rain (0/1) & Both log-linear model and negative binomial model are tested \\
  \cite{Niemeier:1996aa} & Morning flag, rain flag, high temp flag, location variable, season variable & A Poisson bicycle count model is assumed and fitted \\
  \cite{Nosal:2014aa} & Temperature (F), humidity (\%), rain flag, rain prev. 3hrs, am rain, pm rain & The relationship  is analyzed using a log-linear regression model \\
  \cite{Parkin:2008aa} & Gender, car ownership, hilliness, off-road routes proportion & A logistic regression model of relevant socio-economic and physical variables is estimated. \\
  \cite{Pinjari:2009aa} & Household density, employment density, fraction of commercial land area, fraction of senior, fraction of race population, etc. & The model system takes the form of a joint mixed Multinomial Logit–Multiple Discrete-Continuous Extreme Value (MNL–MDCEV) structure \\
  \cite{Rose:2011aa} & Temerature(C), rainfall(mm), holiday flag, school seanson flag, day of the week & Weather and other effects examined using an aggregate model of daily ridership \\
  \cite{Thomas:2009aa} & Temperature(C), sunshine(h), precipitation(mm), wind force(m/s), cycle path use & A bi-level structure is developed with the upper level being a log-linear model and the lower level being a linear model \\
  \bottomrule
\label{tb:lit}
\end{ctabular}
\end{scriptsize}
\end{centering}

@@

\begin{centering}
\begin{table}[ht!]
\begin{scriptsize}
\caption{Variables included in model specification}
\begin{tabularx}{0.5\textwidth}{>{\raggedright\arraybackslash}p{1.6cm}>{\raggedright\arraybackslash}X}
\toprule
Variable & Description \\
\midrule
Count* & Number of bicycles per day \\
Daylight  & Length of daylight in hours \\
UW & Was the University of Washington in session? \\
Holiday & Was the day a holiday as recognized by UW? \\
Max temp & Maximum temperature for the day \\
Max precip & Daily maximum inches of precipitation in any hour \\
Sat--Fri & Day of the week dummy variable (relative to Sunday) \\
Day \# & Sequentially numbered day of study \\
\bottomrule
\multicolumn{2}{r}{* Dependent variable}
\end{tabularx}

@@

While control of the weather and seasons are admittedly beyond the
scope of policy makers, this research does suggest that planners and
policy makers may want to develop strategies that help mitigate the
impacts of the natural environment during the winter months on the
increasing number of bicyclists. Future research could focus on
determining what, if any, programmatic or built interventions could
ameliorate unfavorable cold- and wet-weather bicycling conditions.

>>>>

% Layout options
\chapterstyle{article}
\setulmarginsandblock{.75in}{1in}{*}
\setlrmarginsandblock{.75in}{*}{*}
\setcolsepandrule{.33in}{0pt}

\checkandfixthelayout
\raggedbottom

@@

Bicycling offers a variety of social benefits that include the health
of those choosing to bicycle regularly, as well as, more broadly, a
society-wide increase in mobility when people shift from congestion
causing modes to bicycling. As such, bicycling is being promoted by
policy makers and urban planners as an increasingly important
component of urban transportation. In order to develop policy and
infrastructural improvements to induce more bicycling, a robust
understanding of factors related to bicycling is required.

This article contributes to that understanding by focusing on the
substantive effects of weather and seasonality on the number of
bicyclists observed on a given day. Specifically, we hypothesize that
factors associated with inclement weather (esp. low temperature and
precipitation) will be inversely correlated with bicycle counts. We
similarly expect to see an additional decrease in counts associated
with both physical and social/institutional dimensions of season. That
is, in seasons where the days get shorter, we expect fewer cyclists
due to a kind of decision making inertia. We also expect fewer
cyclists outside the ``school season'' as we believe that a large
number of bicycling trips are associated with commuting to schools
that, unlike most other institutions have clear annual cycles.

@@

\newfixedcaption{\freetabcaption}{table}
\begin{centering}
\begin{scriptsize}
\freetabcaption{Summary of findings from literature review}
\begin{ctabular}{p{.6in}p{1.15in}p{1.15in}}
  \toprule
  Source & Variable(s) identified & Methods \\
  \midrule
  \cite{Fields:2012aa} & Average daily temperature (F); Total weekly precipitation (inches) & Identify patterns through scatter plots; No explicit model is established. \\
  \cite{Gallop:2012aa} & Temperature, Relative humidity, wind speed, clearness,fog, precipitation & Use ARIMA to account for serial correlation patterns \\
  \cite{Griswold:2011aa} & Nearby population and employment density, proximity to downtown/freeway, age, education level, income, et. & Log linear ordinary least squares regression is used to estimate a bicycle count model \\
  \cite{Helbich:2014aa} & Daily maiximum air temperature (C), daily average wind speed (m/s), daily precipitation(mm) & Place-specific associations of weather conditions are explored through geographically weighted logit models \\
  \cite{Hunt:2007aa} & Descriptive variables indicating lane use, secured parking, level of experience, etc. & Logistic model of cycling-related choices \\
  \cite{Jones:2010aa} & Length of bicycle network, employment density, population density & Standard ordinary least squares regression \\
  \cite{Lewin:2011aa} & Max temp (F), rain flag, snow flag, weekend flag, over 90 flag & Standard linear regression model \\
  \cite{McCahil:2008aa} & logarithmic choice measure, population density, worker density & A new space syntax theory is used to evaluate and predict the bicycle volume throughout a network \\
  \cite{Miranda-Moreno:2011aa} & Temperature (C), humidity (\%), rain presence (0/1), rain in prev. 3hrs (0/1), warm \& humid (0/1), morning rain (0/1) & Both log-linear model and negative binomial model are tested \\
  \cite{Niemeier:1996aa} & Morning flag, rain flag, high temp flag, location variable, season variable & A Poisson bicycle count model is assumed and fitted \\
  \cite{Nosal:2014aa} & Temperature (F), humidity (\%), rain flag, rain prev. 3hrs, am rain, pm rain & The relationship  is analyzed using a log-linear regression model \\
  \cite{Parkin:2008aa} & Gender, car ownership, hilliness, off-road routes proportion & A logistic regression model of relevant socio-economic and physical variables is estimated. \\
  \cite{Pinjari:2009aa} & Household density, employment density, fraction of commercial land area, fraction of senior, fraction of race population, etc. & The model system takes the form of a joint mixed Multinomial Logit–Multiple Discrete-Continuous Extreme Value (MNL–MDCEV) structure \\
  \cite{Rose:2011aa} & Temerature(C), rainfall(mm), holiday flag, school seanson flag, day of the week & Weather and other effects examined using an aggregate model of daily ridership \\
  \cite{Thomas:2009aa} & Temperature(C), sunshine(h), precipitation(mm), wind force(m/s), cycle path use & A bi-level structure is developed with the upper level being a log-linear model and the lower level being a linear model \\
  \bottomrule
\label{tb:lit}
\end{ctabular}
\end{scriptsize}
\end{centering}

@@

\begin{centering}
\begin{table}[ht!]
\begin{scriptsize}
\caption{Variables included in model specification}
\begin{tabularx}{0.47\textwidth}{>{\raggedright\arraybackslash}p{1.6cm}>{\raggedright\arraybackslash}X}
\toprule
Variable & Description \\
\midrule
Count* & Number of bicycles per day \\
Daylight  & Length of daylight in hours \\
UW & The University of Washington in session (TRUE/FALSE) \\
Holiday & The day a holiday as recognized by UW (TRUE/FALSE) \\
Max temp & Maximum temperature in Fahrenheit for the day \\
Max precip & Daily maximum inches of precipitation in any hour \\
Sat--Fri & Day of the week dummy variable (TRUE/FALSE, relative to Sunday) \\
Day \# & Sequentially numbered day of study \\
\bottomrule
\multicolumn{2}{r}{* Dependent variable}
\end{tabularx}

@@

While control of the weather and seasons are admittedly beyond the
scope of policy makers, this research does suggest that planners and
policy makers may want to develop strategies that help mitigate the
impacts of the natural environment during the winter months on the
increasing number of bicyclists. In other words, the delta between
warm dry days and cold wet days should be treated as the opportunity
frontier. Future research could focus on determining what, if any,
programmatic or built interventions could ameliorate unfavorable cold-
and wet-weather bicycling conditions.
